---
title: "Comparing Air Pollution Trends Between Delhi, India and Dhaka, Bangladesh from 2003-2023"
subtitle : "ENV 872L Final Project"
author: "Aadya Shukla and Humayra Rahman"
date: "`r Sys.Date()`"
output: pdf_document
toc: true
---

\listoffigures
\listoftables

\newpage

## Introduction
Air pollution is one of the most pressing environmental and public health challenges of the 21st century, particularly across South Asia, where rapid urbanization, industrial growth, vehicular emissions, and agricultural waste burning have made several cities among the most polluted in the world (Ditta et al., 2026). Delhi, the capital of India, and Dhaka, the capital of Bangladesh, are two such megacities that consistently rank among the worst for air quality globally, with both cities experiencing severe seasonal pollution cycles driven by winter temperature inversions, stubble burning, brick kilns, and vehicle emissions along the shared Indo-Gangetic Plain (Ditta et al., 2026; Begum et al., 2018). Despite their geographic proximity and shared pollution sources, the two cities have followed divergent regulatory paths. India launched its National Clean Air Programme in 2019 targeting significant reductions in particulate matter (Garg et al., 2022), while Bangladesh was among the first South Asian nations to ban two-stroke engine vehicles in the early 2000s and transition to CNG-powered alternatives (State of Global Air, n.d.), raising important questions about whether these differing policy trajectories have translated into meaningfully different pollution trends over time.

While prior studies have individually examined air quality trends in Delhi (Garg et al., 2022; Singh et al., 2023) and Dhaka (Islam et al., 2021; Begum & Hopke, 2018), long-term comparative analyses spanning both cities remain limited. This project addresses that gap by using 20 years of monthly air quality data (2003 to 2023) to compare air quality trends in PM2.5, PM10, ozone (O3), and AQI across both cities, testing whether there is a significant seasonal component or decreasing trend, whether Delhi's pollution is significantly worse than Dhaka's, and whether the two cities are improving at statistically different rates.
Thus, this study addresses the following hypotheses:

$H_1$: Air quality in Delhi and Dhaka exhibits a statistically significant seasonal trend over the 2003–2023 study period.

$H_2$: Air quality in Delhi and Dhaka exhibits a statistically significant long-term decreasing trend over the 2003–2023 study period.

$H_3$: Delhi's air quality is significantly worse than Dhaka's, as measured by mean concentrations of PM2.5, PM10, O3, and AQI over the 2003–2023 study period.

## Data description

### Dhaka Dataset
The Dhaka air quality data was sourced from the Bangladesh Air Quality Index (AQI) Dataset (2000–2025), published by Tapon Paul of Daffodil International University and retrieved from Mendeley Data (Paul, 2026). The full dataset contains 1,048,551 hourly records spanning 103 cities across Bangladesh across 13 columns: city_id, city_name, lat, lon, datetime, pm10, pm2_5, carbon_monoxide, carbon_dioxide, nitrogen_dioxide, sulphur_dioxide, ozone, and aqi. AQI values are calculated according to the US EPA standard. 

### Delhi Dataset
The Delhi air quality data was sourced from the Processed Delhi Air Quality and Weather Data dataset curated by Abhinav Sarkar and retrieved from Hugging Face (Sarkar, 2024). The full dataset contains approximately 2.9 million hourly observations from multiple monitoring locations across Delhi from March 2000 to November 2024, across 15 columns: city, event_timestamp, temperature, humidity, pressure, wind_speed, wind_direction, pm25, pm10, no2, so2, o3, co, aqi, and location_id. The dataset was pre-processed prior to publication, with missing AQI values removed and remaining gaps addressed through mean imputation and nearest neighbor interpolation. 

## Methods

### Initial Data Wrangling

Both datasets were cleaned and processed separately before being combined for analysis. For the Dhaka dataset, we filtered the to retain only observations from Dhaka city. For both datasets, the datetime columns were converted into a proper R datetime object using the ymd_hms function and only the columns needed for analysis were selected: Location, Month, Year, PM2.5, PM10, O3, and AQI. Both datasets were filtered to January 1, 2003 through December 31, 2023. We chose this window because the Dhaka dataset spans a longer date range from 2000 to 2025 and the Delhi dataset does not have complete monthly records for 2000 and 2024. To keep the data consistent and comparable across both cities, we focused on a clean 20-year trend. Hourly observations for both datasets were aggregated into monthly averages for all four pollutants (AQI, PM 2.5, PM 10, and O3) using na.rm = TRUE to handle any missing values. A Location variable was added to each dataset before merging and the two datasets were combined using rbind. A date column marking the first day of each month was created using make_date for use in visualizations and regression analyses. The final combined dataset contained 504 monthly observations over 8 columns (see below).


```{r Loading packages, checking working directory, message=FALSE, warning=FALSE, echo=FALSE}

library(dplyr); library(lubridate); library(tidyverse);
library(here); library(zoo); library(trend);
library(Kendall); library(tseries); library(ggplot2)

#here()

```

```{r Delhi Dataset, eval=FALSE, echo=FALSE}

Delhi_dataset_raw <- read.csv(here(
  "DelhiDataset/delhi_air_quality_feature_store_processed.csv"),
  stringsAsFactors = TRUE)

Delhi_dataset_summarized <- Delhi_dataset_raw %>% 
  select(city, event_timestamp, pm25, pm10, o3, aqi) %>% 
  mutate(event_timestamp = ymd_hms(event_timestamp)) %>% 
  filter(event_timestamp >= ymd("2003-01-01") &
    event_timestamp <= ymd("2023-12-31")) %>% 
  mutate(month = month(event_timestamp),
         year = year(event_timestamp)) %>% 
  group_by(month, year) %>% 
  summarize(
    PM2.5 = mean(pm25, na.rm = TRUE),
    PM10 = mean(pm10, na.rm = TRUE),
    O3 = mean(o3, na.rm = TRUE),
    AQI = mean(aqi, na.rm = TRUE)
  ) %>% 
  mutate(Location = "Delhi")

write.csv(Delhi_dataset_summarized, 
          here("processed/delhi_final_processed.csv"), row.names = FALSE)
  
```

```{r Dhaka Dataset, eval=FALSE, echo=FALSE}
#reading csv file
BD_dataset_raw <- read.csv(here("AQI Bangladesh.csv"),
        stringsAsFactors = TRUE)

#filtering for Dhaka only
dhaka_dataset_raw <- BD_dataset_raw %>%
  filter(city_name == "Dhaka")

view(dhaka_dataset_raw)

#filtered for date,location,pm2.5,pm10,aqi,o3
dhaka_air <- dhaka_dataset_raw %>%
  mutate(datetime = ymd_hm(datetime)) %>%
  filter(city_name == "Dhaka") %>%
  select(city_name, datetime, ozone, pm2_5, pm10, aqi) %>%
  rename(
    Location  = city_name,
    Date = datetime,
    O3        = ozone,
    PM2.5     = pm2_5,
    PM10      = pm10,
    AQI       = aqi
  ) 

# Filter date range
dhaka_air <- dhaka_air %>%
  filter(Date >= as.Date("2003-01-01") & Date <= as.Date("2023-12-31"))

dhaka_dataset_summarized <- dhaka_air %>%
  mutate(month = month(Date),
         year  = year(Date)) %>%
  group_by(month, year) %>%
  summarize(
    PM2.5 = mean(PM2.5, na.rm = TRUE),
    PM10  = mean(PM10,  na.rm = TRUE),
    O3    = mean(O3,    na.rm = TRUE),
    AQI   = mean(AQI,   na.rm = TRUE)
  ) %>%
  mutate(Location = "Dhaka")

view(dhaka_dataset_summarized)

write.csv(dhaka_dataset_summarized, 
          here("processed/dhaka_final_processed.csv"), row.names = FALSE)

```

```{r Combining Datasets Into a New Dataframe, warning=FALSE, message=FALSE, echo=FALSE}
#reading both datasets
dhaka_final_processed <- read.csv(here("processed/dhaka_final_processed.csv"))
delhi_final_processed <- read.csv(here("processed/delhi_final_processed.csv"))

#combining into one dataframe
combined_dataset <- rbind(dhaka_final_processed, delhi_final_processed)

combined_dataset <- mutate(combined_dataset,
                           first.month = make_date(year, month, 1))

```

```{r warning=FALSE, message=FALSE, echo=FALSE}

str(combined_dataset)

#colSums(is.na(combined_dataset))
```

### Visualization

To get an initial sense of how pollution has changed over time in each city, we plotted monthly trends for all four pollutants using ggplot2. Both cities were displayed on the same graph, with a linear trend line added using geom_smooth(method = lm) to highlight the overall direction of change across the 20-year study period.

### Time Series

We built monthly time series objects for each pollutant and each city using the ts() function, starting in January 2003 with a frequency of 12. Each series was then decomposed using Seasonal and Trend decomposition using Loess (STL) via the stl() function with s.window = "periodic". STL breaks each time series into three components: a seasonal cycle, a long-term trend, and a remainder. This was particularly useful for our analysis because air pollution in both cities follows a strong annual seasonal pattern, and we needed to isolate the underlying long-term trend from that recurring noise to address Hypothesis 1.

### Trend Analysis

To test whether pollution levels were changing significantly over time, we ran both seasonal and non-seasonal Mann-Kendall tests for each pollutant in each city. We used the Seasonal Mann-Kendall test (smk.test() from the trend package) as our primary test because it is designed for data with seasonal patterns. It compares each month only to the same month in other years, which avoids confounding the trend with seasonal variation. We also ran the non-seasonal Mann-Kendall test (MannKendall() from the Kendall package) on the de-seasonalized series, which we obtained by subtracting the seasonal component from the STL decomposition. Comparing the two sets of results helped us confirm whether any detected trends were robust.

### Comparative Analysis

To compare air quality between the two cities, we used two complementary approaches. First, we ran two-sample t-tests for each pollutant to test whether mean pollution levels over the full study period were significantly different between Delhi and Dhaka, addressing Hypothesis 2. Second, we fitted linear regression models with an interaction term between time and location for each pollutant, using the formula lm(pollutant ~ first.month * Location). The interaction term first.month:LocationDhaka was the key result we were looking for. A significant interaction would mean that the two cities were changing at statistically different rates over time, directly addressing Hypothesis 3.


```{r Checking for normality AQI}
par(mfrow = c(1, 2))
qqnorm(combined_dataset$AQI[combined_dataset$Location == "Delhi"],
       main = "Delhi AQI")
qqline(combined_dataset$AQI[combined_dataset$Location == "Delhi"],
       col = "blue")
qqnorm(combined_dataset$AQI[combined_dataset$Location == "Dhaka"],
       main = "Dhaka AQI")
qqline(combined_dataset$AQI[combined_dataset$Location == "Dhaka"],
       col = "blue")
```

```{r Checking for normality PM 2.5}
par(mfrow = c(1, 2))
qqnorm(combined_dataset$PM2.5[combined_dataset$Location == "Delhi"],
       main = "Delhi PM 2.5")
qqnorm(combined_dataset$PM2.5[combined_dataset$Location == "Dhaka"],
       main = "Dhaka PM2.5")
```

```{r Checking for normality PM 10}
par(mfrow = c(1, 2))
qqnorm(combined_dataset$PM10[combined_dataset$Location == "Delhi"],
       main = "Delhi PM 10")
qqnorm(combined_dataset$PM10[combined_dataset$Location == "Dhaka"],
       main = "Dhaka PM 10")
```

```{r Checking for normality Ozone}
par(mfrow = c(1, 2))
qqnorm(combined_dataset$O3[combined_dataset$Location == "Delhi"],
       main = "Delhi O3")
qqnorm(combined_dataset$O3[combined_dataset$Location == "Dhaka"],
       main = "Dhaka O3")
```

## Results

```{r fig.cap = "Monthly AQI Trend from 2003-2023", echo=FALSE, message=FALSE, warning=FALSE}

ggplot(combined_dataset, aes(y = AQI, x = first.month, colour = Location)) + 
  geom_line() +
    theme_classic() +
    scale_color_manual(values = c("Delhi" = "#1abc9c", "Dhaka" = "purple")) +
    theme(legend.background = element_rect(colour = "black", fill = "white",
    linewidth = 0.5), plot.title = element_text(hjust = 0.5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  geom_smooth(method = lm) +
  labs(
    x = "Date", 
    y = "AQI",
    title = "Average Monthly AQI Trend in Delhi and Dhaka\n(2003-2023)",
    colour = "Locations")

```

```{r fig.cap = "Monthly PM 2.5 Trend from 2003-2023", echo=FALSE, message=FALSE, warning=FALSE}

ggplot(combined_dataset, aes(y = PM2.5, x = first.month, colour = Location)) + 
  geom_line() +
    theme_classic() +
    scale_color_manual(values = c("Delhi" = "#1abc9c", "Dhaka" = "purple")) +
    theme(legend.background = element_rect(colour = "black", fill = "white",
    linewidth = 0.5), plot.title = element_text(hjust = 0.5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  geom_smooth(method = lm) +
  labs(
    x = "Date",
    y = expression(paste("PM 2.5 ", , "(", mu, "g/", m^3, ")")),
    title = "Average Monthly PM 2.5 Trend in Delhi and Dhaka\n(2003-2023)",
    colour = "Locations")

```

```{r fig.cap = "Monthly PM 10 Trend from 2003-2023", echo=FALSE, message=FALSE, warning=FALSE}

ggplot(combined_dataset, aes(y = PM10, x = first.month, colour = Location)) + 
  geom_line() +
    theme_classic() +
    scale_color_manual(values = c("Delhi" = "#1abc9c", "Dhaka" = "purple")) +
    theme(legend.background = element_rect(colour = "black", fill = "white",
    linewidth = 0.5), plot.title = element_text(hjust = 0.5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  geom_smooth(method = lm) +
  labs(
    x = "Date",
    y = expression(paste("PM 10 ", , "(", mu, "g/", m^3, ")")),
    title = "Average Monthly PM 10 Trend in Delhi and Dhaka\n(2003-2023)",
    colour = "Locations")

```

```{r fig.cap = "Monthly Ozone Trend from 2003-2023", echo=FALSE, message=FALSE, warning=FALSE}

ggplot(combined_dataset, aes(y = O3, x = first.month, colour = Location)) + 
  geom_line() +
    theme_classic() +
    scale_color_manual(values = c("Delhi" = "#1abc9c", "Dhaka" = "purple")) +
    theme(legend.background = element_rect(colour = "black", fill = "white",
    linewidth = 0.5), plot.title = element_text(hjust = 0.5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  geom_smooth(method = lm) +
  labs(
    x = "Date",
    y = expression(paste("Ozone ", , "(", mu, "g/", m^3, ")")),
    title = "Average Monthly Ozone Trend in Delhi and Dhaka\n(2003-2023)",
    colour = "Locations")

```

## Discussion
## Limitations
## Conclusion
## References

## TIME SERIES ANALYSES

```{r Creating time series objects}
# Monthly AQI for Delhi
Delhi.AQI.monthly.ts <- ts(delhi_final_processed$AQI,
                           start = c(2003,1), frequency = 12)
Delhi.AQI.monthly.decomposed <- stl(Delhi.AQI.monthly.ts,
                                    s.window = "periodic")
plot(Delhi.AQI.monthly.decomposed)

# Monthly Ozone for Delhi
Delhi.Ozone.monthly.ts <- ts(delhi_final_processed$O3,
                             start = c(2003,1), frequency = 12)
Delhi.Ozone.monthly.decomposed <- stl(Delhi.Ozone.monthly.ts,
                                    s.window = "periodic")
plot(Delhi.Ozone.monthly.decomposed)


# Monthly PM2.5 for Delhi
Delhi.PM2.5.monthly.ts <- ts(delhi_final_processed$PM2.5,
                             start = c(2003,1), frequency = 12)
Delhi.PM2.5.monthly.decomposed <- stl(Delhi.PM2.5.monthly.ts,
                                    s.window = "periodic")
plot(Delhi.PM2.5.monthly.decomposed)


# Monthly PM10 for Delhi
Delhi.PM10.monthly.ts <- ts(delhi_final_processed$PM10,
                            start = c(2003,1), frequency = 12)
Delhi.PM10.monthly.decomposed <- stl(Delhi.PM10.monthly.ts,
                                    s.window = "periodic")
plot(Delhi.PM10.monthly.decomposed)



# Monthly AQI for Dhaka
Dhaka.AQI.monthly.ts <- ts(dhaka_final_processed$AQI,
                           start = c(2003,1), frequency = 12)
Dhaka.AQI.monthly.decomposed <- stl(Dhaka.AQI.monthly.ts,
                                    s.window = "periodic")
plot(Dhaka.AQI.monthly.decomposed)


# Monthly Ozone for Dhaka
Dhaka.Ozone.monthly.ts <- ts(dhaka_final_processed$O3,
                             start = c(2003,1), frequency = 12)
Dhaka.Ozone.monthly.decomposed <- stl(Dhaka.Ozone.monthly.ts,
                                    s.window = "periodic")
plot(Dhaka.Ozone.monthly.decomposed)


# Monthly PM2.5 for Dhaka
Dhaka.PM2.5.monthly.ts <- ts(dhaka_final_processed$PM2.5,
                             start = c(2003,1), frequency = 12)
Dhaka.PM2.5.monthly.decomposed <- stl(Dhaka.PM2.5.monthly.ts,
                                    s.window = "periodic")
plot(Dhaka.PM2.5.monthly.decomposed)


# Monthly PM10 for Dhaka
Dhaka.PM10.monthly.ts <- ts(dhaka_final_processed$PM10,
                            start = c(2003,1), frequency = 12)
Dhaka.PM10.monthly.decomposed <- stl(Dhaka.PM10.monthly.ts,
                                    s.window = "periodic")
plot(Dhaka.PM10.monthly.decomposed)

```

```{r Mann-Kendall Tests}


# Seasonal Mann-Kendall Dhaka

# Dhaka AQI
Dhaka.AQI.monthly.trend <- trend::smk.test(Dhaka.AQI.monthly.ts)
Dhaka.AQI.monthly.trend
summary(Dhaka.AQI.monthly.trend)
# Dhaka AQI not significant (p = 0.0768)


# Dhaka PM2.5
Dhaka.PM2.5.monthly.trend <- trend::smk.test(Dhaka.PM2.5.monthly.ts)
Dhaka.PM2.5.monthly.trend
summary(Dhaka.PM2.5.monthly.trend)
# Dhaka PM2.5 not significant (p = 0.0768)

# Dhaka PM10
Dhaka.PM10.monthly.trend <- trend::smk.test(Dhaka.PM10.monthly.ts)
Dhaka.PM10.monthly.trend
summary(Dhaka.PM10.monthly.trend)
# Dhaka PM2.5 not significant (p = 0.1068)

# Dhaka O3
Dhaka.Ozone.monthly.trend <- trend::smk.test(Dhaka.Ozone.monthly.ts)
Dhaka.Ozone.monthly.trend
summary(Dhaka.Ozone.monthly.trend)
# Dhaka ozone (O3) is significant (1.432e-06)

# Seasonal Mann-Kendall Delhi

# Delhi AQI
Delhi.AQI.monthly.trend <- trend::smk.test(Delhi.AQI.monthly.ts)
Delhi.AQI.monthly.trend
summary(Delhi.AQI.monthly.trend)
# Delhi AQI not significant (0.3226)

# Delhi PM2.5
Delhi.PM2.5.monthly.trend <- trend::smk.test(Delhi.PM2.5.monthly.ts)
Delhi.PM2.5.monthly.trend
summary(Delhi.PM2.5.monthly.trend)
# Delhi AQI not significant (0.08624)

# Delhi PM10
Delhi.PM10.monthly.trend <- trend::smk.test(Delhi.PM10.monthly.ts)
Delhi.PM10.monthly.trend
summary(Delhi.PM10.monthly.trend)
# Delhi PM10 not significant (0.4257)

# Delhi O3
Delhi.Ozone.monthly.trend <- trend::smk.test(Delhi.Ozone.monthly.ts)
Delhi.Ozone.monthly.trend
summary(Delhi.Ozone.monthly.trend)
# Delhi ozone (O3) not significant (0.3098)


# Non-Seasonal Mann-Kendall Dhaka

# Dhaka AQI
Dhaka.monthly.AQI.nonseasonal <- Dhaka.AQI.monthly.ts - 
  Dhaka.AQI.monthly.decomposed$time.series[,1]

Dhaka.nonseasonal.AQI.trend <- Kendall::MannKendall(
  Dhaka.monthly.AQI.nonseasonal)
Dhaka.nonseasonal.AQI.trend
summary(Dhaka.nonseasonal.AQI.trend)

# Dhaka PM2.5
Dhaka.monthly.PM2.5.nonseasonal <- Dhaka.PM2.5.monthly.ts - 
  Dhaka.PM2.5.monthly.decomposed$time.series[,1]

Dhaka.nonseasonal.PM2.5.trend <- Kendall::MannKendall(
  Dhaka.monthly.PM2.5.nonseasonal)
Dhaka.nonseasonal.PM2.5.trend
summary(Dhaka.nonseasonal.PM2.5.trend)

# Dhaka PM10
Dhaka.monthly.PM10.nonseasonal <- Dhaka.PM10.monthly.ts - 
  Dhaka.PM10.monthly.decomposed$time.series[,1]

Dhaka.nonseasonal.PM10.trend <- Kendall::MannKendall(
  Dhaka.monthly.PM10.nonseasonal)
Dhaka.nonseasonal.PM10.trend
summary(Dhaka.nonseasonal.PM10.trend)

# Dhaka O3
Dhaka.monthly.Ozone.nonseasonal <- Dhaka.Ozone.monthly.ts - 
  Dhaka.Ozone.monthly.decomposed$time.series[,1]

Dhaka.nonseasonal.Ozone.trend <- Kendall::MannKendall(
  Dhaka.monthly.Ozone.nonseasonal)
Dhaka.nonseasonal.Ozone.trend
summary(Dhaka.nonseasonal.Ozone.trend)

# Non-Seasonal Mann-Kendall Delhi

# Delhi AQI
Delhi.monthly.AQI.nonseasonal <- Delhi.AQI.monthly.ts - 
  Delhi.AQI.monthly.decomposed$time.series[,1]

Delhi.nonseasonal.AQI.trend <- Kendall::MannKendall(
  Delhi.monthly.AQI.nonseasonal)
Delhi.nonseasonal.AQI.trend
summary(Delhi.nonseasonal.AQI.trend)

# Delhi PM2.5
Delhi.monthly.PM2.5.nonseasonal <- Delhi.PM2.5.monthly.ts - 
  Delhi.PM2.5.monthly.decomposed$time.series[,1]

Delhi.nonseasonal.PM2.5.trend <- Kendall::MannKendall(
  Delhi.monthly.PM2.5.nonseasonal)
Delhi.nonseasonal.PM2.5.trend
summary(Delhi.nonseasonal.PM2.5.trend)

# Delhi PM10
Delhi.monthly.PM10.nonseasonal <- Delhi.PM10.monthly.ts - 
  Delhi.PM10.monthly.decomposed$time.series[,1]

Delhi.nonseasonal.PM10.trend <- Kendall::MannKendall(
  Delhi.monthly.PM10.nonseasonal)
Delhi.nonseasonal.PM10.trend
summary(Delhi.nonseasonal.PM10.trend)

# Delhi O3
Delhi.monthly.Ozone.nonseasonal <- Delhi.Ozone.monthly.ts - 
  Delhi.Ozone.monthly.decomposed$time.series[,1]

Delhi.nonseasonal.Ozone.trend <- Kendall::MannKendall(
  Delhi.monthly.Ozone.nonseasonal)
Delhi.nonseasonal.Ozone.trend
summary(Delhi.nonseasonal.Ozone.trend)

```

Table: Summary of Mann-Kendall Results

Location | Variable | Seasonal p-value | Non-Seasonal p-value | Trend Statistic
:---: | :---: | :---: |:---: |:---:
Dhaka | AQI | 0.0768 | 0.2045 | -204
Dhaka | PM 2.5 | 0.0768 | 0.2354 | -204
Dhaka | PM 10 | 0.1068 | 0.2354 | -186
Dhaka | O3 | **1.43e-06** | **2.06e-06** | -554
Delhi | AQI | 0.3226 | 0.1078 | -114
Delhi | PM 2.5 | 0.0862 | **0.0330** | -197
Delhi | PM 10 | 0.4257 | 0.1409 | -92
Delhi | O3 | 0.3098 | 0.4421 | 117

```{r two-sample t tests}

# AQI
t.test(AQI ~ Location, data = combined_dataset)

# PM2.5
t.test(PM2.5 ~ Location, data = combined_dataset)

# PM10
t.test(PM10 ~ Location, data = combined_dataset)

# O3
t.test(O3 ~ Location, data = combined_dataset)

```

Table: Summary of two sample t-test Results

Variable | p-value | Dhaka mean | Delhi mean
:---: | :---: | :---: |:---: 
AQI | **2.2e-16** | 170.35 | 275.71
PM2.5 | **2.2e-16**  | 100.21 | 240.54
PM10 | **2.2e-16**  | 169.18 | 242.17
O3| **2.2e-16**  | 50.87 | 31.78



```{r linear regressions}

#Linear Regression

# AQI
AQI.model <- lm(AQI ~ first.month * Location, data = combined_dataset)
summary(AQI.model)

# PM2.5
PM2.5.model <- lm(PM2.5 ~ first.month * Location, data = combined_dataset)
summary(PM2.5.model)

# PM10
PM10.model <- lm(PM10 ~ first.month * Location, data = combined_dataset)
summary(PM10.model)


# O3
O3.model <- lm(O3 ~ first.month * Location, data = combined_dataset)
summary(O3.model)
```

Table: Summary of Linear Regression Results

Variable | p-value | Min | Median | Max | R-squared
:---: | :---: | :---: |:---: |:---: |:---: 
AQI | **2.2e-16** | -152.894 | -1.296 | 93.331 | 0.812
PM2.5 | **2.2e-16** | -127.439 | -3.511 | 79.986 | 0.894
PM10 | **2.2e-16** | -214.902 | -5.708 | 128.747 | 0.634
O3 | **2.2e-16** | -30.487 | -4.640 | 32.740 | 0.424